---
title: "Lecture 6: Examples of Bayesian Networks and Markov Networks"
author:
- name: Zhenke Wu <br>
        Department of Biostatistics, University of Michigan
date: September 22, 2016
header-includes:
- \usepackage{amsmath}
runtime: shiny
output: ioslides_presentation
widescreen: true
---

## Lecture 5 Main Points Once Again
* Bayesian network ($\mathcal{G},P$)
    + Directed acyclic graph (DAG): $\mathcal{G}$, comprised of nodes $V$ and edges $E$
    + Joint distribution $P$ over $|V|$ random variables
    + $P$ is Markov to $\mathcal{G}$ **if** variables in $P$ satisfy $X_A\perp X_B \mid X_C$ whenever $C$ d-separates $A$ and $B$ as read off from $\mathcal{G}$
* Markov network ($\mathcal{H}$,$P$)
    + Undirected graph (UG): $\mathcal{H}$, comprised of nodes $V$ and edges $E$
    + Joint distribution $P$ over $|V|$ random variables
    + $P$ is Global Markov to $\mathcal{H}$ **if** variables in $P$
    satisfy $X_A\perp X_B \mid X_C$ whenever $C$ separates $A$ and $B$ as read off from the graph
* Roughly, given Markov properties, graph $\mathcal{G}$, or $\mathcal{H}$ is a valid guide to understand the variable relationships in distribution $P$






## Lecture 5 Main Points Once Again (continued)
* **Question:** Given a distribution $P$ that is Markov to a DAG $\mathcal{G}$, can we find an UG $\mathcal{H}$ with the same set of nodes
so that $P$ is also Markov to it? (Yes, by **moralization**---"marrying the parents". But UG could lose some d-separations, e.g., v-structure; won't lose any if $\mathcal{G}$ is already moralized.)
* (Question above, but with DAG and UG reserved) (Yes, by constructing directed edges following certain node ordering. But DAG could lose some separations, e.g., four-node loop)
* Are there distributions representable by both DAG and UG, but without loss of (d-)separations? (Yes.) If so, under what conditions? (Those distributions that
are Markov to a **chordal network**: any non-adjacent pair of nodes in a loop must be connected.)



## Markov Network Example: Ising Model

* A mathematical model of ferromagnetism in statistical mechanics; Named after physicist Ernst Ising; 
* The model consists of discrete variables that represent magnetic dipole moments of atomic spins that can be in one of two states (+1 or −1). 
* The spins are arranged in a graph, usually a lattice, allowing each spin to interact with its neighbors.


## Markov Network Example: Ising Model

* **Formulation**: Let $\mathcal{H}=(V,E)$ be an undirected graph, e.g., (lattice or non-lattice). Let the binary random variables $X_{i}\in\{-1,+1\}$. The Ising model takes the form

\[
P(\mathbf{x}; \theta) \propto \exp \left(\sum_{i\in V}\theta_ix_i + \sum_{(i,j)\in E} \theta_{ij}x_ix_j\right)
\]

* From the model form, Ising model is positive and Markov to $\mathcal{H}$. Using the local Markov property, and code the $-1$ into $0$, the conditional distribution for a node $X_i$ given all its neighbors is given by a logisitic regression:

\[
Pr(X_i=1 \mid X_j, j\neq i; \mathbf{\theta})
= Pr(X_i=1 \mid X_j, (i,j)\in E; \mathbf{\theta})
\]

\[
= sigmoid(\theta_i +\sum_{j:(i,j)\in E}\theta_{ij}x_j)
\]

## Markov Network Example: Special case of Ising Model 

* No external field: $\theta_i=0, X_i\in V$
* $\theta_{ij}=\beta J$, $\forall i,j$. 
* We have 

\[
P(\mathbf{x}; \mathbf{\theta}) \propto \left( \beta\cdot J\cdot\sum_{(i,j)\in E} x_ix_j\right)
\]

* $\beta$: inverse temperature; large $\beta$, lower temperature (colder)
* $J>0$: neighboring nodes tend to align, so-called ferromagnetic model; $J<0$: anti-ferromagnetic.


## Square-Lattice Ising Model under Different Temperatures {.columns-2}
* $P(\mathbf{x}; \theta) \propto \exp \left( \beta\cdot J\cdot\sum_{(i,j)\in E} x_ix_j\right)$
    - Set $J=2$, ferromagnetic
    - Vary inverse temperature: $\beta$
    - Try different graph size: $n^2$
```{r, echo=FALSE,fig.wdith=5,fig.height=5}
## Gibbs sampler for the Ising model

f.gibbs <- function(x,beta)
{
  ## Purpose: Gibbs sampler for the Ising model. It alternates between
  ##          updating values at positions (i,j) with i+j even and
  ##          updating values at positions where i + j odd.
  ## -------------------------------------------------------------------------
  ## Arguments: x=configuration (n times n matrix with n even and values +1/-1
  ##            beta=inverse temperature
  ## -------------------------------------------------------------------------
  ## Author: Hans-Ruedi Kuensch, Date: 22 Nov 2001, 12:10
  beta <- 2*beta
  n <- nrow(x)
  nq <- 0.25*n*n # Number of indices divided by 4.
  ger <- 2*(1:(n/2)) # even indices
  ung <- ger-1  # odd indices 
  z <- rep(0,n) # set x to zero outside the {1,2,...,n}^2
  nb <- rbind(x[-1,],z) + rbind(z,x[-n,]) + cbind(x[,-1],z) + cbind(z,x[,-n])
  # sum of neighbors
  nb1 <- nb[ung,ung] # i,j both odd
  pr <- exp(beta*nb1) # Compute conditional probabilities given the neighbors
  pr <- pr/(pr+1)
  x[ung,ung] <- 2*(pr > runif(nq))-1 # update of x
  nb1 <- nb[ger,ger] # update for i,j both even, same thing
  pr <- exp(beta*nb1)
  pr <- pr/(pr+1)
  x[ger,ger] <- 2*(pr > runif(nq))-1
  nb <- rbind(x[-1,],z) + rbind(z,x[-n,]) + cbind(x[,-1],z) + cbind(z,x[,-n])
  # recomputing the sum of neighbors
  nb1 <- nb[ung,ger] # update for i odd, j even
  pr <- exp(beta*nb1)
  pr <- pr/(pr+1)
  x[ung,ger] <- 2*(pr > runif(nq))-1
  nb1 <- nb[ger,ung] # update for i even, j odd
  pr <- exp(beta*nb1)
  pr <- pr/(pr+1)
  x[ger,ung] <- 2*(pr > runif(nq))-1
  x
}

inputPanel(
  sliderInput("n", label = h1("n: grid points"),
              min = 20, max = 300, value = 32,step = 10),
  sliderInput("beta", label = h1("beta: inverse-temperature"),
              min = 0, max = 0.5, value = 0.1)
)
```


```{r,echo=FALSE}
renderPlot({
  beta <- input$beta
  n <- input$n #be patient
  x <- matrix(2*(runif(n*n) < 0.5)-1,nrow=n,ncol=n) #Startwert i.i.d.
  image(1:n,1:n,z=x,col=gray(0:1),xaxs="i",yaxs="i",pty="s",
        axes=F,xlab="",ylab="")
  for (i in (1:1000)) x <- f.gibbs(x,beta)
  image(1:n,1:n,z=x,col=gray(0:1),xaxs="i",yaxs="i",pty="s",
        axes=F,xlab="",ylab="")
  mtext(side = 3,"Simulated Spatial Pattern",cex=3,line = 2)
},height=500,width=500)

```


## Bayesian Network Example: Naive Bayes for SPAM classification {.columns-2}
```{r fig.width=5, fig.height=4,echo=FALSE}
library(png)
library(grid)
img <- readPNG("spam.png")
 grid.raster(img)
```

* Features (words) assumed **independent** given SPAM or HAM status, hence "naive"
* Infer the SPAM status given observed evidence from the email
* Very fast, low storage requirements, robust to irrelevant features, good for benchmarking


## Bayesian Network Example: Beta-Binomial Model {.columns-2}

```{r fig.width=5, fig.height=4,echo=FALSE}
library(png)
library(grid)
img <- readPNG("score_rate.png")
 grid.raster(img)
```

* 30 soccer players' penalty shot score rates and the actual number of shots
* What's the best estimtate of a player's scoring rate? (empirial Bayes estimate)
* Information from other players could contribute to a given player's score rate estimate. Use moralized graph to explain.

## Inference for Bayesian Network: Moralization
* **Question:** given observed evidence, what's the updated probability distribution for those unobserved variables? Or more specifically, which conditional independencies still hold, which don't?
* **Proposition 4.7**
    - Let $\mathcal{G}$ be a Bayesian Network over $\mathbf{V}$ and $\mathbf{Z}=\mathbf{z}$ an observation. Let $\mathbf{W}=\mathbf{V}-\mathbf{Z}$. Then $P_{\mathcal{G}}(\mathbf{W}\mid \mathbf{Z}=\mathbf{z})$ is a Gibbs distribution defined by factors $\Phi=\{\phi_{X_i}\}_{X_i\in \mathbf{V}}$, where $\phi_{X_i} = P_{\mathcal{G}}(X_i\mid Pa_{X_i})[\mathbf{Z}=\mathbf{z}].$
The partition function for this Gibbs distribution is $P_{\mathcal{G}}(\mathbf{Z}=\mathbf{z})$, the marginal probability.
* Use the moralized graph to identify conditional independencies given observed data. 
* Because the above Gibbs distribution factorizes according to a moralized graph $M(\mathcal{G})$ which creates cliques for a family (parents and a child). 
* And factorization of $P$ with respect to $M(\mathcal{G})$ is equivalent to $P$ satisfying the Markov property, which means you can use the moralized graph as a "map".


## Moralized Graph
* Natually, if a Bayesian network is already moral (parents are connected by directed edges), then moralization will not add extra edges and conditional independencies will not be lost. 
* So in this case separations in UG $M({\mathcal{G}})$ correspond one-to-one for d-separations in the original DAG $\mathcal{G}$.


## Chordal Graph
* If $\mathcal{H}$ is an UG, and let $\mathcal{G}$ be any DAG that is minimal I-map for $\mathcal{H}$, then $\mathcal{G}$ must have no immoralities. [Proof]
* Nonchordal DAGs must have immoralities
* $\mathcal{G}$ then must be chordal

* The conditional independencies encoded by an undirected chordal graph can be perfectly encoded by a directed graph. (Use clique tree proof)
* If $\mathcal{H}$ is nonchordal, no DAG can encode **perfectly** the same set of conditional independencies as in $\mathcal{H}$. (Use the third bullet point.)

## The connections among graphs and distributions (note from Lafferty, Liu and Wasserman) {.columns-2}

```{r fig.width=5, fig.height=4,echo=FALSE}
library(png)
library(grid)
img <- readPNG("intersection.png")
 grid.raster(img)
```


* The intersection of Bayesian networks and Markov networks (or random fields) are those distributions Markov to a chordal graph
* Chordal DAGs mean the absence of immoralities
* Chordal graph $\Leftrightarrow$ decomposable graph

## Comment

* **Next Lecture**: Overview of Module 2 that discusses inference: more algorithmic-flavored and exciting ideas. Begin exact inference.

* **No required reading**.

* **Homework 1** due 11:59PM, October 3rd, 2016 to Instructor's email.





